---
title: "Lab 10: Distances and the Border Zone"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: Samantha Nauman
    email: snauman4@colostate.edu
format: html
---
```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)

# Data
library(AOI)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
library(dplyr)
```
# Question 1:

## 1.1 Define a Projection
```{r}
# North America Equidistant Conic:
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```
## 1.2 Get USA State Boundaries
```{r}
remotes::install_github("mikejohnson51/AOI", force = TRUE)

# getting USA state boundaries
state_bounds <- aoi_get(state = 'conus')

# putting data for a projected coordinate system
eqdc_state_bounds <- (eqdc_state_bounds = st_transform(state_bounds, eqdc)) %>%
  filter(!state_abbr %in% c("HI","AK","PR"))
```
## 1.3 Get country boundaries for Mexico, USA, and Canada
```{r}
country_bounds <- aoi_get(country = c("MX", "CA", "USA"))

eqdc_country_bounds <- (eqdc_country_bounds = st_transform(country_bounds, eqdc))
```
## 1.4 Get city locations from the CSV file
```{r}
us_cities <- readr::read_csv("data/uscities.csv")

# converting the data frame into a spatial object
us_cities_df <- data.frame(
  city = us_cities$city,
  state = us_cities$state_name,
  state_id = us_cities$state_id,
  population = us_cities$population,
  X = us_cities$lng,
  Y = us_cities$lat) %>%
  filter(!state %in% c("Hawaii","Alaska","Puerto Rico"))
  # using st_as_sf
usc_gcs <- st_as_sf(us_cities_df, 
                    coords = c("X", "Y"), 
                    crs = 4326)
  # remove cities in unwanted states and put the data in a pcd (st_transform)
eqdc_cities <- st_transform(usc_gcs, crs = eqdc) 
st_transform(eqdc_cities, eqdc)
```
# Question 2

## 2.1 Distance to USA Border (coastline or national) (km)
```{r}
us_border <- eqdc_state_bounds %>%
  st_union() %>%
  st_cast("MULTILINESTRING") 

eqdc_cities$distance_to_border <- st_distance(eqdc_cities, us_border)

eqdc_cities <- eqdc_cities %>%
  mutate(distance_to_border = as.numeric(distance_to_border[,1]/1000))

top5_farthest <- eqdc_cities %>%
  slice_max(order_by = distance_to_border, n = 5, with_ties = FALSE) %>%
  st_drop_geometry() %>%
  mutate(distance_to_border = paste0(round(distance_to_border,1), "km"))

flextable::flextable(top5_farthest)
```
## 2.2 Distance to States (km)
```{r}
state_borders_preserved <- eqdc_state_bounds %>%
  st_cast("MULTILINESTRING")

dist_matrix_state <- st_distance(eqdc_cities, state_borders_preserved)

eqdc_cities <- eqdc_cities %>%
  mutate(dist_to_state_border_km = as.numeric(dist_matrix_state[,1])/1000)

farthest_cities_from_states <- eqdc_cities %>%
  arrange(desc(dist_to_state_border_km)) %>%
  slice(1:5) %>%
  st_drop_geometry() %>%
  select(city_name = city, state, dist_to_state_border_km) %>%
  mutate(dist_to_state_border_km = paste0(round(dist_to_state_border_km, 1), "km"))

flextable::flextable(farthest_cities_from_states)
```
## 2.3 Distance to Mexico (km)
```{r}
# isolate Mexico
Mexico <- eqdc_country_bounds %>%
  filter(sovereignt == "Mexico")

Mexico <- st_transform(Mexico, eqdc)

mx_border <- Mexico %>%
  st_cast("MULTILINESTRING")

# calculate distance 
dist_matrix_mexico <- st_distance(eqdc_cities, mx_border)

eqdc_cities <- eqdc_cities %>%
  mutate(distance_to_mexico = as.numeric(dist_matrix_mexico[,1])/1000)

top5_mx <- eqdc_cities %>%
  arrange(desc(distance_to_mexico)) %>%
  slice(1:5) %>%
  st_drop_geometry() %>%
  select(city, state, distance_to_mexico)

flextable::flextable(top5_mx)
```
## 2.4 Distance to Canada (km)
```{r}
# isolate Canada
Canada <- eqdc_country_bounds %>%
  filter(sovereignt == "Canada")

Canada <- st_transform(Canada, eqdc)

cd_border <- Canada %>%
  st_cast("MULTILINESTRING")

# calculate distance 
dist_matrix_canada <- st_distance(eqdc_cities, cd_border)

eqdc_cities <- eqdc_cities %>%
  mutate(distance_to_canada = as.numeric(dist_matrix_canada[,1])/1000)

top5_cd <- eqdc_cities %>%
  arrange(desc(distance_to_canada)) %>%
  slice(1:5) %>%
  st_drop_geometry() %>%
  select(city, state, distance_to_canada)

flextable::flextable(top5_cd)
```
# Question 3

## 3.1 Data
```{r}
top10_cities <- eqdc_cities |>
  arrange(desc(population)) |>
  slice_head(n=10)

ggplot() +
  geom_sf(data = eqdc_country_bounds, fill = "grey", color = "black", lty = 1) +
  geom_sf(data = eqdc_state_bounds, fill = NA, color = "grey", lty = 2, size = 0.5) +
  geom_sf(data = us_border, fill = NA, color = "black", size = 0.7) +
  geom_sf(data = top10_cities, color = "blue", size = 2) + 
  ggrepel::geom_label_repel(
    data = top10_cities,
    aes(geometry = geometry, label = city), 
    stat = "sf_coordinates",
    size = 3) + 
  theme_minimal()
```
## 3.2 City Distance from the Border
```{r}
# Cities farthest from US border is: top5_farthest
top5_us_border <- eqdc_cities |>
  arrange(desc(distance_to_border)) |>
  slice_head(n=5)

ggplot() +
  geom_sf(data = us_border, fill = NA, color = "black", lty = 1) +
  geom_sf(data = eqdc_state_bounds, fill = NA, color = "grey", lty = 2, size = 0.5) +
  geom_sf(data = eqdc_cities, aes(color = dist_to_state_border_km), size = 1) +
  scale_color_viridis_c(option = "magma", name = "Distance to US Border (km)") + 
  geom_sf(data = top5_us_border, color = "blue", size = 2) + 
  ggrepel::geom_label_repel(
    data = top5_us_border,
    aes(geometry = geometry, label = city), 
    stat = "sf_coordinates",
    size = 3) + 
  theme_minimal()
```
## 3.3 City Distance from Nearest State
```{r}
top5_state_border <- eqdc_cities |>
  arrange(desc(dist_to_state_border_km)) |>
  slice_head(n=5)

ggplot() +
  geom_sf(data = us_border, fill = NA, color = "black", lty = 1) +
  geom_sf(data = eqdc_state_bounds, fill = NA, color = "grey", lty = 2, size = 0.5) +
  geom_sf(data = eqdc_cities, aes(color = dist_to_state_border_km), size = 1) +
  scale_color_viridis_c(option = "magma", name = "Distance to State Borders (km)") + 
  geom_sf(data = top5_state_border, color = "blue", size = 2) + 
  ggrepel::geom_label_repel(
    data = top5_state_border,
    aes(geometry = geometry, label = city), 
    stat = "sf_coordinates",
    size = 3) + 
  theme_minimal()
```
## 3.4 Equidistance boundary from Mexico and Canada
```{r}
# Create absolute difference column 
eqdc_cities <- eqdc_cities |>
  mutate(abs_diff_can_mex = abs(distance_to_mexico - distance_to_canada))

dist_matrix_mexico <- st_distance(eqdc_cities, mx_border)
  
eqdc_cities <- eqdc_cities |> 
  mutate(distance_to_mexico = as.numeric(dist_matrix_mexico[,1])/1000)

# Filter cities within +/- 100km of equidistance
equidistant_cities <- eqdc_cities |>
  filter(abs_diff_can_mex <= 100)

# Get 5 most populous equidistant cities
top5_eqdist <- equidistant_cities |>
  arrange(desc(population)) |>
  slice_head(n=5)

# Plot
ggplot() +
  geom_sf(data = us_border, fill = NA, color = "black") +
  geom_sf(data = eqdc_state_bounds, fill = NA, color = "grey", size = 0.5) +
  geom_sf(data = equidistant_cities, aes(color = abs_diff_can_mex), size = 1) +
  scale_color_viridis_c(name = "|Canada - Mexico| Distance (km)") +
  geom_sf(data = top5_eqdist, color = "red", size = 2) +
  gghighlight(abs_diff_can_mex <= 100, use_direct_label = FALSE) +  
  ggrepel::geom_label_repel(
    data = top5_eqdist,
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3
  ) +
  theme_minimal() +
  labs(
    title = "Cities ~100 km Equidistant from Canadian and Mexican Borders",
    subtitle = "Top 5 Most Populous Cities Highlighted"
  )
```
# Question 4

## 4.1 Quantifying Border Zone
```{r}
cities_in_border_zone <- eqdc_cities |>
  filter(distance_to_border <= 160)

num_cities_in_border_zone <- nrow(cities_in_border_zone)

total_population_border_zone <- sum(cities_in_border_zone$population)

total_population_all_cities <- sum(eqdc_cities$population)

percent_population_in_border_zone <- (total_population_border_zone / total_population_all_cities)*100

# Create summary table
border_summary_tbl <- tibble(
  `Number of Cities` = num_cities_in_border_zone,
  `Total Population in Border Zone` = total_population_border_zone,
  `Total Population (All Cities)` = total_population_all_cities,
  `Percent Population in Border Zone` = round(percent_population_in_border_zone, 1)
)

# Convert to flextable
library(flextable)
flextable(border_summary_tbl) |>
  set_caption("Summary of Cities within 160 km of U.S. National Border") |>
  colformat_int(j = c("Number of Cities", "Total Population in Border Zone", "Total Population (All Cities)")) |>
  colformat_num(j = "Percent Population in Border Zone", suffix = "%", digits = 1) |>
  autofit() |>
  theme_vanilla()
```
## 4.2 Mapping Border Zone
```{r}
top10_borderzone <- cities_in_border_zone |>
  arrange(desc(population)) |>
  slice_head(n=10)

library(gghighlight)

ggplot() + 
  geom_sf(data = eqdc_cities, aes(color = distance_to_border), size = 1) +  
  geom_sf(data = us_border, fill = NA, color = "black") +
  geom_sf(data = eqdc_state_bounds, fill = NA, color = "grey", size = 0.5) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Distance to Border (km)") +
  geom_sf(data = top10_borderzone, color = "blue", size = 2) +
  gghighlight(distance_to_border <= 100, use_direct_label = FALSE) +
  ggrepel::geom_label_repel(
    data = top10_borderzone,
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    box.padding = 0.3,
    max.overlaps = Inf
  ) +
  labs(title = "Cities Within 100-Mile Border Zone",
       subtitle = "Top 10 most populous cities highlighted",
       caption = "Distance from national border (km)") +
  theme_minimal()
```
## 4.3 Instead of labeling the 10 most populous cities, label the most populous city in each state within the Danger Zone
```{r}
most_populous_per_state <- cities_in_border_zone |>
  group_by(state) |>
  slice_max(order_by = population, n=1, with_ties = FALSE)

ggplot() +
  geom_sf(data = eqdc_cities, aes(color = distance_to_border), size = 1) +  
  geom_sf(data = us_border, fill = NA, color = "black") +
  geom_sf(data = eqdc_state_bounds, fill = NA, color = "grey", size = 0.5) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Distance to Border (km)") +
  geom_sf(data = most_populous_per_state, color = "blue", size = 2) +
  gghighlight(distance_to_border <= 100, use_direct_label = FALSE) +
  ggrepel::geom_label_repel(
    data = most_populous_per_state,
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    box.padding = 0.3,
    max.overlaps = Inf
  ) +
  labs(title = "Most Populous Cities Within Their State in the 100-Mile Border Zone",
       subtitle = "Most populous cities highlighted",
       caption = "Distance from national border (km)") +
  theme_minimal()
```